Understanding meta-analysis through data simulation with applications to power analysis

### 2023-07-18 documents/paper/paper.tex
fixing after revision

  \usepackage{unicode-math}

@@

@@

\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}

@@

\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}

@@

\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}

@@

\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}

@@

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{meta-analysis, monte carlo simulations, power analysis\newline\indent Word count: 7877}
\usepackage{csquotes}
\usepackage{setspace}
\captionsetup[figure]{font={stretch=1,footnotesize}}
\usepackage{float}
\usepackage{orcidlink}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Understanding meta-analysis through data simulation with applications to power analysis},
  pdfauthor={Filippo Gambarota 1 \& Gianmarco Altoè 1},
  pdflang={en-EN},
  pdfkeywords={meta-analysis, monte carlo simulations, power analysis},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

\abstract{%
Meta-analysis is a powerful tool to combine evidence from existing literature. Despite several introductory and advanced materials about organizing, conducting, and reporting a meta-analysis, to our knowledge, there are no introductive materials about simulating the most common meta-analysis models. Data simulation is essential for developing and validating new statistical models and procedures. Furthermore, data simulation is a powerful educational tool for understanding a statistical method. In this tutorial, we show how to simulate fixed-effects, random-effects, and meta-regression models and illustrate how to estimate statistical power. In the supplementary materials, we extended these simulations also for multilevel and multivariate models. All materials associated with this article can be accessed on Open Science Framework (\url{https://osf.io/54djn/}).
}

@@

A meta-analysis is an essential tool for combining knowledge from multiple studies quantitatively. Meta-analysis is commonly used together with a systematic review of the literature. The meta-analysis has several advantages. Firstly, it allows combining evidence from multiple studies assigning more weight to studies with lower estimation variability. Then using meta-regression, it is possible to assess the impact of study-level variables that could impact the phenomenon. Finally, considering the replication crisis, there are statistical methods to determine the presence and extent of the publication bias. Despite the advantages, meta-analysis implementation is not always straightforward, especially for complex data structures. Additionally, the fact that there are several introductory and advanced resources to understand meta-analysis (Borenstein, Hedges, Higgins, \& Rothstein, 2009; Harrer, Cuijpers, Furukawa, \& Ebert, 2021; Schmid, Stijnen, \& White, 2022), to our knowledge, there are no introductory resources about how to simulate realistic meta-analytic data. Simulating data has several advantages because it requires understanding the statistical method and the data-generation process. Furthermore, data simulation is the primary tool when it comes to evaluating a new analysis method, estimating the statistical power, or understanding the long-run behavior of our data generation process (Gelman \& Hill, 2006, pp. 155--176; Gelman, Hill, \& Vehtari, 2020, pp. 69--76; Ingalls, 2011). A recent paper by DeBruine and Barr (2021), which deeply inspired the current work, proposed a stimulating way to understand linear mixed effects via data simulation. Simulating data is also a powerful educational tool within this framework. For these reasons, this work aims to introduce the basic concepts of meta-analysis and Monte Carlo simulations for fixed and random-effects, and meta-regression models with applications also for statistical power calculation. In the first section, we will introduce basic concepts of the meta-analysis that are useful for setting up the simulation. We will evaluate the effect size, variance calculation, and the fixed vs.~random effects model distinction. Then we will describe how to simulate data for these models and simulate a meta-regression model with a categorical and numerical predictor. Finally, we will introduce the power analysis extending the previous examples to estimate the statistical power. We used the R statistical programming language (version 4.2.3; R Core Team, 2023). We assume the reader is familiar with the basic concepts of R, but core functions will be explained. Code and materials are available on the OSF repository (\url{https://osf.io/54djn/}). Multivariate and multilevel models' simulation examples and more details about the coding approach are available in the supplementary materials.

@@

As an example that will be used throughout the paper, we consider the efficacy of memory training in improving memory performance during a cognitive task. The typical primary study will collect data from a group of participants receiving the memory training (\emph{experimental group}) and another group receiving a control treatment (\emph{control group}). The focus of the meta-analysis is collecting multiple studies with similar aims and methods and estimating the average effect of memory training. Despite differences in the type of cognitive task or experimental setup, each primary study collects an \emph{experimental group} (\(n_{C}\)) and \emph{control group} (\(n_{C}\)) and computes the average performance (\emph{experimental group} \({\bar y}_{T}\) and \emph{control group} \({\bar y}_{C}\)) and standard deviations (\emph{experimental group} \(s_{T}\) and \emph{control group}
\(s_{C}\)).

@@

The first step of a meta-analysis is to standardize information from each included study to convert each effect into a standard measure. This common measure should give an immediate idea of the direction (i.e., the treatment improves or reduces performance) and the size of the effect. Standardized effect size measures (see Lakens, 2013 for an overview), such as the Standardized Mean Difference (SMD), which could be estimated by Cohen's \(d\) (Cohen, 1988) or the Pearson correlation coefficient \(\rho\) which could be calculated using the associated sample estimator \emph{r} are commonly used to compare heterogeneous outcome variables. If all studies used the same raw measure, such as reaction times, it is possible to directly meta-analyze the studies without standardizing. In practice, studies often use heterogeneous measures, therefore requiring standardized measures.

As reported in the previous section, beyond the effect size of each study, we need to assign a weight according to the precision. For this reason, we need to calculate the sampling variability of the effect size or the raw measure that will represent the estimation precision. Each raw or standardized effect size measure (e.g., raw mean difference, Cohen's \(d\), and Pearson's correlation) has a different formula to calculate the sampling variability. The idea is to choose the appropriate measure considering the study design (e.g., between vs.~within-subjects) and available information and find the proper formula to compute the sampling variability. In addition, there are formulas and approaches to convert from one effect size measure to another (Borenstein et al., 2009; Lakens, 2013; Lipsey \& Wilson, 2001).\footnote{See also the blog post by James Pustejovsky (\href{https://www.jepusto.com/alternative-formulas-for-the-smd/}{link}) for a general method to compute the sampling variability for standardized mean difference measures.} Usually, the effect size sampling variability depends mainly on the sample size that determines the weight assigned during the meta-analytic estimation.

\hypertarget{fixed-vs-random-effects-model}{%
\subsubsection{Fixed vs random-effects model}\label{fixed-vs-random-effects-model}}

The core of a meta-analysis is combining the results of multiple studies giving more weight to studies that provide a more precise effect estimation. We can divide meta-analysis models into two categories: the \emph{fixed-effects} and the \emph{random-effects model}. If we imagine combining multiple studies replicating the same experiment, we assume the real treatment effect \(\theta_{f}\) is \emph{fixed,} and each study is a more or less precise estimation of this parameter. This is an essential and often inappropriate assumption because it ignores that study-level characteristics may impact the actual treatment effect.

Suppose some study-level characteristics (e.g., participants' age, sex, or socioeconomic status) or the experimental paradigm (e.g., type of memory task or difficulty) could impact the treatment effect. In that case, the assumption of the \emph{fixed-effects model} is no longer plausible. There is no unique real effect but a distribution of effects with mean \(\theta_{r}\) and variance \(\tau^{2}\). The mean is the \emph{average} effect, and the variance is the heterogeneity of the phenomenon. The \emph{random-effects model} will estimate the mean (\(\theta_{r}\)) and the heterogeneity \(\tau^{2}\) will help generalize to the population of effects beyond the pool of included studies.

Most importantly, the parameters estimated with the \emph{fixed} and \emph{random-effects} do not represent the same quantity. The \emph{random-effects model} estimates the \emph{average} effect across a distribution of true effects, while the \emph{fixed-effects model} estimates a single true effect. Figure 1 depicts the idea of the fixed vs.~random-effects model. The \emph{fixed-effects model} can be considered a particular type of \emph{random-effects model} where \(\tau^{2} = 0\) thus, the average effect is also the unique true effect.

@@

{\centering \includegraphics[width=0.8\linewidth]{paper_files/figure-latex/img-fixed-vs-random-1}

@@

\caption{The difference between the assumptions of the \emph{fixed} and \emph{random-effects model}. Each distribution depicts the sampling distribution of \(k = 4\) hypothetical studies (\(i = 1\), \(2\), \ldots, \(4\)) with a certain observed effect size \(d_{i}\), sampling variability \(\sigma_{i}^{2}\) and the 95\% confidence interval (black segment). The \emph{fixed-effects} plot on the left suggests that each observed effect size has the same underlying true effect \(\theta_{f}\) (each distribution has the same mean) with a different degree of precision (e.g., Study 1 is more precise than Study 3). In practice, each study has a different observed effect size where studies with high precision (i.e., narrow sampling distributions) will be remarkably close to the real effect (\(\theta_{f}\)). The \emph{random-effects model} on the right suggests that beyond the error term (\(\epsilon_{i}\)), each real effect size is composed of a fixed part (now \(\theta_{r}\)) and a random part (\(\theta_{i}\), black segments below the sampling distributions). The random part is determined by \(\tau^{2}\), which represents the variability of the effect size distribution at the population level. When \(\tau^{2}\) is zero, the random-effects reduces to a fixed-effects model.}\label{fig:img-fixed-vs-random}
\end{figure}

\hypertarget{multilevel-and-multivariate-models}{%
\subsubsection{Multilevel and multivariate models}\label{multilevel-and-multivariate-models}}

The most straightforward situation for a meta-analysis is when each included study contributes with a single effect size. However, it is common to have multiple effect sizes belonging to the same study creating a situation of dependency between statistical units. There are numerous sources of dependency for a meta-analysis (Cheung, 2014; 2015, pp. 121--122). The first type of dependency (\emph{multilevel}) consists of multiple effect sizes from independent participants. Despite being collected with different participants, effect sizes from the same study could be correlated. We obtain a multilevel data structure with independent effect sizes in the same study. There are multiple ways to model this data structure (see Cheung, 2014 for an overview). Still, the most common way is to use a \emph{multilevel meta-analysis model,} considering the heterogeneity between and within studies. The standard \emph{random-effects} (but also the \emph{fixed-effects model}) model is also called a \emph{two-level} model because it combines data after aggregating the participants-level (\emph{first-level}) data. The \emph{three-level} model is commonly used to manage the multilevel data structure (Cheung, 2014, 2019). Compared to the \emph{two-level} model, the \emph{three-level} model estimates two heterogeneity parameters (\(\tau^{2}\) and \(\omega^{2}\)), one for the heterogeneity between studies and the other for the heterogeneity within studies.

A second type of dependency (\emph{multivariate}) consists of multiple effect sizes collected from the same pool of participants. For example, when using two cognitive tasks on the same pool of participants. This is commonly known as \emph{multivariate meta-analysis} because the effect sizes sampling errors are correlated given that participants are assessed multiple times. A \emph{multivariate meta-analysis model} could be used to
consider the correlation between sampling errors.

A third type of dependency arises when true effect sizes are correlated at the population level (Cheung, 2015, pp. 121--122). For example, two different memory tasks could be correlated because the latent psychological constructs are intrinsically correlated. This correlation is present even when other groups of participants perform the two memory tasks (i.e., no \emph{multivariate} dependency) or are administered by different researchers (i.e., no \emph{multilevel} dependency). This type of dependency could be managed by estimating the correlation between outcomes using a \emph{multivariate random-effects model} (Cheung, 2015, pp. 127--130). Figures \ref{fig:img-multilevel} and \ref{fig:img-multivariate} depict the conceptual idea of \emph{multilevel} and \emph{multivariate} models.

{\centering \includegraphics[width=0.8\linewidth]{img/multilevel}

\caption{Graphical representation of a \emph{multilevel} (three-level) model. On the top, the effect sizes are distributed at the population level. Then, each study is defined as the overall effect and the random effects determined by \(\tau^{2}\), \(\theta_{r} + \theta_{i}\). Finally, each experiment (effect size) within a study is composed of the study-specific effect \(\theta_{r} + \theta_{i}\) and the experiment-specific random-effects \(\theta_{r} + \theta_{i} + \theta_{\text{ij}}\) determined by \(\omega^{2}\).}\label{fig:img-multilevel}
\end{figure}

{\centering \includegraphics[width=0.8\linewidth]{img/multivariate.pdf }

\caption{Graphical representation of a \emph{multivariate} model. In this example, we have two outcomes (magenta and green) with an average effect, heterogeneity parameters, and a certain true correlation value (\(\rho\), third type of dependency). Each study at level-2 (as in the standard random-effects model) has two effect sizes with corresponding sampling variances. The two effect sizes are correlated (\(\rho_{s}\), with \(s\) for sampling errors) because each effect is collected on the same participants.}\label{fig:img-multivariate}
\end{figure}

@@

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{set.seed}\NormalTok{(}\DecValTok{123}\NormalTok{) }\CommentTok{\# seed for simulation to ensure the reproducibility of results}
\NormalTok{theta }\OtherTok{\textless{}{-}} \FloatTok{0.5} \CommentTok{\# real mean difference in the populations}
\NormalTok{sigma }\OtherTok{\textless{}{-}} \DecValTok{1} \CommentTok{\# real standard deviations in the population}
\NormalTok{n }\OtherTok{\textless{}{-}} \DecValTok{30} \CommentTok{\# sample size for both groups}
\NormalTok{nsims }\OtherTok{\textless{}{-}} \FloatTok{1e5} \CommentTok{\# number of simulations = 100000}

\CommentTok{\# simulates the sampling distribution of the mean difference}
\NormalTok{d\_mc }\OtherTok{\textless{}{-}} \FunctionTok{replicate}\NormalTok{(nsims, }\AttributeTok{expr =}\NormalTok{ \{}
\NormalTok{  g1 }\OtherTok{\textless{}{-}} \FunctionTok{rnorm}\NormalTok{(}\AttributeTok{n =}\NormalTok{ n, }\AttributeTok{mean =}\NormalTok{ theta, }\AttributeTok{sd =}\NormalTok{ sigma)}
\NormalTok{  g2 }\OtherTok{\textless{}{-}} \FunctionTok{rnorm}\NormalTok{(}\AttributeTok{n =}\NormalTok{ n, }\AttributeTok{mean =} \DecValTok{0}\NormalTok{, }\AttributeTok{sd =}\NormalTok{ sigma)}
  \FunctionTok{mean}\NormalTok{(g1) }\SpecialCharTok{{-}} \FunctionTok{mean}\NormalTok{(g2) }\CommentTok{\# calculates the sample mean difference and returns the value}
\NormalTok{\})}

@@

We load relevant packages and set the \texttt{seed} for reproducibility to prepare the R environment for the simulations.

@@

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{library}\NormalTok{(dplyr) }\CommentTok{\# for data manipulation}
\FunctionTok{library}\NormalTok{(tidyr) }\CommentTok{\# for data manipulation}
\FunctionTok{library}\NormalTok{(metafor) }\CommentTok{\# for meta{-}analysis models fitting}
\NormalTok{seed }\OtherTok{\textless{}{-}} \DecValTok{2023} \CommentTok{\# general seed for all simulations}
\end{Highlighting}
\end{Shaded}

@@

Before diving into the specific simulations, in this section, we define the common aspects of all simulations in the following sections. The current paper focuses on the \emph{two-level}, \emph{fixed,} and \emph{random-effects model}s. All the examples refer to primary studies that assess the efficacy of a treatment by comparing a control and an experimental group. Equation \eqref{eq:datagen} formalizes the assumption about generating participants' data within a single study. The two groups (\emph{experimental} \(y_{T}\) and \emph{control} \(y_{C}\)) are generated from normal distributions with \(\mu = 0\) for the control group and \(\mu = \theta\) (where \(\theta\) is the effect size) for the experimental group. Both groups are assumed to be sampled from populations with homogeneous variances \(\sigma^{2} = 1\).

\begin{align}
\begin{gathered}
y_{Ti} \sim \mathcal{N}(\theta, 1) \\
y_{Ci} \sim \mathcal{N}(0, 1)
\label{eq:datagen}
\end{gathered}
\end{align}

The true Standardized Mean Difference can be estimated using Cohen's \(d\) as reported in Equation \eqref{eq:effsize}. The difference between the two means is divided by the pooled standard deviation of the two groups. It is common to apply a small-sample correction version of this effect size called Hedges' \(g\) (Hedges, 1981)

\begin{align}
\begin{gathered}
d = \frac{\bar y_T - \bar y_C}{S_p} \\
S_p = \sqrt{\frac{S^2_{y_T}(n_{y_T} - 1) + S^2_{y_C}(n_{C} - 1)}{n_{T} + n_{C} - 2}} \\
\sigma^2_d = \frac{n_{T} + n_{C}}{n_{T}n_{C}} + \frac{d^2}{2(n_{T} + n_{C} - 2)}
\label{eq:effsize}
\end{gathered}
\end{align}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Choose a \(\theta\), \(n_{T}\), and \(n_{C}\) value.
\item
  Simulate \(n_{T}\) observations from a Gaussian distribution with \(\mu = \theta\) and \(\sigma^{2} = 1\) and \(n_{T}\) observations from a Gaussian distribution with \(\mu = 0\) and \(\sigma^{2} = 1\). This way, the expected difference between groups will be \(d\) and the variance is fixed at 1.
\item
  Calculate the pooled standard deviation \(S_{p}\) that is an estimation of the population standard deviation (\(\sigma\)) to standardize the mean difference.
\item
  Calculate the observed effect size \(d\) and the sampling variance \(\sigma_{d}^{2}\).
\end{enumerate}

@@

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{set.seed}\NormalTok{(seed)}
\FunctionTok{sim\_study}\NormalTok{(}\AttributeTok{theta =} \FloatTok{0.3}\NormalTok{, }\AttributeTok{nc =} \DecValTok{10000}\NormalTok{, }\AttributeTok{nt =} \DecValTok{10000}\NormalTok{, }\AttributeTok{aggregate =} \ConstantTok{FALSE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

>>>>

  \usepackage{unicode-math} % this also loads fontspec

@@

\ifPDFTeX\else
  % xetex/luatex font selection

@@

\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}

@@

\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}

@@

\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}

@@

\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}

@@

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{meta-analysis, monte carlo simulations, power analysis\newline\indent Word count: 7877}
\usepackage{csquotes}
\usepackage{setspace}
\captionsetup[figure]{font={stretch=1,footnotesize}}
\usepackage{float}
\usepackage{orcidlink}
\newcommand{\rev}[1]{\textcolor{red}{#1}}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Understanding meta-analysis through data simulation with applications to power analysis},
  pdfauthor={Filippo Gambarota 1 \& Gianmarco Altoè 1},
  pdflang={en-EN},
  pdfkeywords={meta-analysis, monte carlo simulations, power analysis},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

\abstract{%
Meta-analysis is a powerful tool to combine evidence from existing literature. Despite several introductory and advanced materials about organizing, conducting, and reporting a meta-analysis, to our knowledge, there are no introductive materials about simulating the most common meta-analysis models. Data simulation is essential for developing and validating new statistical models and procedures. Furthermore, data simulation is a powerful educational tool for understanding a statistical method. In this tutorial, we show how to simulate \rev{equal-effects}, random-effects, and meta-regression models and illustrate how to estimate statistical power. In the supplementary materials, we extended these simulations also for multilevel and multivariate models. All materials associated with this article can be accessed on Open Science Framework (\url{https://osf.io/54djn/}).
}

@@

A meta-analysis is an essential tool for combining knowledge from multiple studies quantitatively. Meta-analysis is commonly used together with a systematic review of the literature. The meta-analysis has several advantages. Firstly, it allows combining evidence from multiple studies assigning more weight to studies with lower estimation variability. \color{red} Then using meta-regression, it is possible to include variables (i.e., moderators) to explain the observed heterogeneity (Borenstein, Hedges, Higgins, \& Rothstein, 2009, pp. 187--203). More recently, location-scale models have been developed to include predictors also on the residual heterogeneity (i.e., Viechtbauer \& López-López, 2022). \color{black} Finally, considering the replication crisis, there are statistical methods to determine the presence and extent of the publication bias. Despite the advantages, meta-analysis implementation is not always straightforward, especially for complex data structures. Additionally, the fact that there are several introductory and advanced resources to understand meta-analysis (Borenstein et al., 2009; Harrer, Cuijpers, Furukawa, \& Ebert, 2021; Schmid, Stijnen, \& White, 2022), to our knowledge, there are no introductory resources about how to simulate realistic meta-analytic data. Simulating data has several advantages because it requires understanding the statistical method and the data-generation process. Furthermore, data simulation is the primary tool when it comes to evaluating a new analysis method, estimating the statistical power, or understanding the long-run behavior of our data generation process (Gelman \& Hill, 2006, pp. 155--176; Gelman, Hill, \& Vehtari, 2020, pp. 69--76; Ingalls, 2011). A recent paper by DeBruine and Barr (2021), which deeply inspired the current work, proposed a stimulating way to understand linear mixed effects via data simulation. Simulating data is also a powerful educational tool within this framework. For these reasons, this work aims to introduce the basic concepts of meta-analysis and Monte Carlo simulations for \textcolor{red}{equal} and random-effects, and meta-regression models with applications also for statistical power calculation. In the first section, we will introduce basic concepts of the meta-analysis that are useful for setting up the simulation. We will evaluate the effect size, variance calculation, and the \textcolor{red}{equal} vs.~random effects model distinction. Then we will describe how to simulate data for these models and simulate a meta-regression model with a categorical and numerical predictor. Finally, we will introduce the power analysis extending the previous examples to estimate the statistical power. We used the R statistical programming language (version 4.2.3; R Core Team, 2023). We assume the reader is familiar with the basic concepts of R, but core functions will be explained. Code and materials are available on the OSF repository (\url{https://osf.io/54djn/}). \color{red} A theoretical introduction, simulation examples for multivariate and multilevel models, and more details about the coding approach are available in the supplementary materials. \color{black}

@@

\color{red} As an example that will be used throughout the paper, we consider the efficacy of memory training in improving memory performance during a cognitive task. The typical primary study will collect data from a group of participants receiving the memory training (\emph{experimental group}) and another group receiving a control treatment (\emph{control group}). The focus of the meta-analysis is collecting multiple studies with similar aims and methods and estimating the average effect of memory training. Despite differences in the type of cognitive task or experimental setup, each primary study collects an \emph{experimental group} (\(n_{T}\)) and \emph{control group} (\(n_{C}\)) and computes the average performance (\emph{experimental group} \(\overline T\) and \emph{control group} \(\overline C\)) and standard deviations (\emph{experimental group} \(s_{T}\) and \emph{control group}
\(s_{C}\)). \color{black}

@@

The first step of a meta-analysis is to extract information from included studies. This common measure should give an immediate idea of the direction (i.e., the treatment improves or reduces performance) and the size of the effect. Standardized effect size measures (see Lakens, 2013 for an overview), such as the Standardized Mean Difference (SMD), which could be estimated by Cohen's \(d\) (Cohen, 1988) or the Pearson correlation coefficient \(\rho\) which could be calculated using the associated sample estimator \emph{r} are commonly used to compare heterogeneous outcome variables\footnote{\color{red} When using bounded effect size indexes such as the Pearson correlation, Fisher's \emph{z} transformation is commonly used (Borenstein et al., 2009, pp. 41--43) \color{black}}. If all studies used the same raw measure, such as reaction times, it is possible to directly meta-analyze the studies without standardizing e.g.~using the unstandardized mean difference (UMD, Borenstein et al., 2009, pp. 21--24).

As reported in the previous section, beyond the effect size of each study, we need to assign a weight according to the precision. For this reason, we need to calculate the sampling variability of the effect size or the raw measure that will represent the estimation precision. Each raw or standardized effect size measure (e.g., raw mean difference, Cohen's \(d\), and Pearson's correlation) has a different formula to calculate the sampling variability. The idea is to choose the appropriate measure considering the study design (e.g., between vs.~within-subjects) and available information and find the proper formula to compute the sampling variability. In addition, there are formulas and approaches to convert from one effect size measure to another (Borenstein et al., 2009; Lakens, 2013; Lipsey \& Wilson, 2001).\footnote{See also the blog post by James Pustejovsky (\href{https://www.jepusto.com/alternative-formulas-for-the-smd/}{www.jepusto.com/alternative-formulas-for-the-smd}) for a general method to compute the sampling variability for standardized mean difference measures.} Usually, the effect size sampling variability depends mainly on the sample size that determines the weight assigned during the meta-analytic estimation.

\hypertarget{equal-vs-fixed-vs-random-effects-model}{%
\subsubsection{Equal vs fixed vs random-effects model}\label{equal-vs-fixed-vs-random-effects-model}}

\color{red}

The core of a meta-analysis is combining the results of multiple studies giving more weight to studies that provide a more precise effect estimation. We can find essentially three meta-analysis models: the equal-effects, the fixed-effects and the random-effects model (Hedges \& Vevea, 1998; Laird \& Mosteller, 1990). The \emph{equal-effects} model assume that each study included in the meta-analysis is a more or less precise estimation of the true underlying effect (\(\theta\)). In other terms, we are assuming that there is no variability (i.e., heterogeneity) among true effect sizes. On the other side, suppose some study-level characteristics (e.g., participants' age, sex, or socioeconomic status) or the experimental paradigm (e.g., type of memory task or difficulty) could impact the treatment effect. In this case we have true variability (i.e., heterogeneity) among studies. The \emph{random-effects} model assume a distribution of real effects with mean \(\mu_{\theta}\) and variance \(\tau^{2}\) thus estimating the heterogeneity. Finally, the \emph{fixed-effects} estimate the average effect of the included pool of studies ignoring the presence of heterogeneity. While the distinction between the equal and fixed-effects model is theoretical, the estimated model is the exactly the same\footnote{See \href{https://wviechtb.github.io/metafor/reference/misc-models.html}{wviechtb.github.io/metafor/reference/misc-models.html} for a detailed explanation}. The \emph{random-effects} model assume and estimate heterogeneity among effect sizes leading to different model parameters. From an inferential point of view, the \emph{random-effects} model provide unconditional inference to the population of effect sizes while the \emph{fixed-effects} model estimate the average effect of the selected studies, and not population-level parameters. The \emph{equal-effects} model assume a single true population effect to be estimated. The Figure \ref{fig:img-equal-vs-random} depicts the theoretical distinction between the equal and random-effects model. In the presence of heterogeneity, we can include moderators (e.g., the type of experimental setup) into a meta-regression model to explain the effect sizes heterogeneity. \color{red} An interesting proposal to combine evidence from these different models into a single analysis is called Bayesian model-averaged meta-analysis (Berkhout, Haaf, Gronau, Heck, \& Wagenmakers, 2023; Gronau, Heck, Berkhout, Haaf, \& Wagenmakers, 2021). \color{black}

\color{black}

@@

{\centering \includegraphics[width=0.8\linewidth]{paper_files/figure-latex/img-equal-vs-random-1}

@@

\caption{\color{red} The difference between the assumptions of the \emph{equal} and \emph{random-effects model}. Each distribution depicts the sampling distribution of \(k = 4\) hypothetical studies (\(i = 1\), \(2\), \ldots, \(4\)) with a certain observed effect size \(y_{i}\) (red squares), sampling variability \(\sigma_{\epsilon_i}^{2}\) and the 95\% confidence interval (black segment). The \emph{equal-effects} plot on the left suggests that each observed effect size has the same underlying true effect \(\theta\) (each distribution has the same mean) with a different degree of precision (e.g., Study 1 is more precise than Study 3). In practice, each study has a different observed effect size where studies with high precision (i.e., narrow sampling distributions) will be close to the real effect (\(\theta\)). The \emph{random-effects model} on the right suggests that beyond the error term (\(\epsilon_{i}\)), each real effect size is composed of a fixed part (now \(\mu_{\theta}\)) and a random part (\(\theta_{i}\)) sampled from a normal distribution with mean zero and variance \(\tau^{2}\). When \(\tau^{2}\) is zero, the random-effects reduces to a equal-effects model. \color{black}}\label{fig:img-equal-vs-random}
\end{figure}

@@

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{set.seed}\NormalTok{(}\DecValTok{123}\NormalTok{) }\CommentTok{\# seed for simulation to ensure the reproducibility of results}
\NormalTok{es }\OtherTok{\textless{}{-}} \FloatTok{0.5} \CommentTok{\# real mean difference in the populations}
\NormalTok{sigma }\OtherTok{\textless{}{-}} \DecValTok{1} \CommentTok{\# real standard deviations in the population}
\NormalTok{n }\OtherTok{\textless{}{-}} \DecValTok{30} \CommentTok{\# sample size for both groups}
\NormalTok{nsims }\OtherTok{\textless{}{-}} \FloatTok{1e5} \CommentTok{\# number of simulations = 100000}

\CommentTok{\# simulates the sampling distribution of the mean difference}
\NormalTok{d\_mc }\OtherTok{\textless{}{-}} \FunctionTok{replicate}\NormalTok{(nsims, }\AttributeTok{expr =}\NormalTok{ \{}
\NormalTok{  g1 }\OtherTok{\textless{}{-}} \FunctionTok{rnorm}\NormalTok{(}\AttributeTok{n =}\NormalTok{ n, }\AttributeTok{mean =}\NormalTok{ es, }\AttributeTok{sd =}\NormalTok{ sigma)}
\NormalTok{  g2 }\OtherTok{\textless{}{-}} \FunctionTok{rnorm}\NormalTok{(}\AttributeTok{n =}\NormalTok{ n, }\AttributeTok{mean =} \DecValTok{0}\NormalTok{, }\AttributeTok{sd =}\NormalTok{ sigma)}
  \FunctionTok{mean}\NormalTok{(g1) }\SpecialCharTok{{-}} \FunctionTok{mean}\NormalTok{(g2) }\CommentTok{\# calculates the sample mean difference and returns the value}
\NormalTok{\})}

@@

\color{red}

This tutorial use several R packages for the simulations, meta-analysis fitting and figures/tables. For the data manipulation we used the \emph{tidyverse} (Wickham, 2023) package. For the models fitting we used the \emph{metafor} (Viechtbauer, 2010) package. For figures and tables the \emph{ggplot2} (Wickham et al., 2023) or the \texttt{metafor::forest()} function, \emph{kableExtra} (Zhu, 2021) and \emph{papaja} (Aust \& Barth, 2022) packages. We set the \texttt{seed} for reproducibility of the simulation environment.
\color{black}

@@

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{library}\NormalTok{(dplyr) }\CommentTok{\# for data manipulation (within tidyverse)}
\FunctionTok{library}\NormalTok{(tidyr) }\CommentTok{\# for data manipulation (within tidyverse)}
\FunctionTok{library}\NormalTok{(metafor) }\CommentTok{\# for meta{-}analysis models fitting}
\FunctionTok{library}\NormalTok{(purrr) }\CommentTok{\# for *apply like functions}
\NormalTok{seed }\OtherTok{\textless{}{-}} \DecValTok{2023} \CommentTok{\# general seed for all simulations}
\end{Highlighting}
\end{Shaded}

@@

\color{red}

Before diving into the specific simulations, in this section, we define the common aspects of all simulations in the following sections. The current paper focuses on the \emph{two-level}, \emph{equal,} and \emph{random-effects} models. All the examples refer to primary studies that assess the efficacy of a treatment by comparing a control and an experimental group. In simulation studies where the purpose is not evaluating effect size estimators, it is convenient to simulate unstandardized effect size measures (see Viechtbauer, 2005, 2007). The estimator is unbiased thus not requiring small sample correction (Hedges, 1981; e.g., Hedges, 1989). Furthermore, the effect size and the sampling variance are independent. Similar to the simulation approach by Viechtbauer (2005, 2007) the \emph{experimental} group (\(T\)) and \emph{control} groups \(C\) are sampled from normal distributions respectively \(T_i \sim \mathcal{N}(\Delta, 1)\) and \(C_i \sim \mathcal{N}(0, 1)\) where \(\Delta\) is the unstandardized mean difference (UMD). The UMD and the sampling variance are calculated using Equations \eqref{eq:effsize} and \eqref{eq:effsizevar} where \(y\) is the estimated value of \(\Delta\). \color{black}

\color{red}

\begin{align}
\begin{gathered}
D = \overline{T} - \overline{C}
\label{eq:effsize}
\end{gathered}
\end{align}

\begin{align}
\begin{gathered}
\sigma_{\epsilon D}^{2} = \frac{s^2_T}{n_T} + \frac{s^2_C}{n_C}
\label{eq:effsizevar}
\end{gathered}
\end{align}

\color{black}

@@

\color{red}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Choose a \(\Delta\), \(n_{T}\), and \(n_{C}\) value.
\item
  Simulate \(n_{T}\) observations from a Gaussian distribution with \(\mu = \Delta\) and \(\sigma^{2} = 1\) and \(n_{T}\) observations from a Gaussian distribution with \(\mu = 0\) and \(\sigma^{2} = 1\). In this way, the expected difference between groups will be \(\Delta\) and the expected variance for each group is 1.
\item
  Calculate the observed effect size \(y\) and the sampling variance \(\sigma_{\epsilon}^{2}\).
\end{enumerate}

The simulation approach can be easily extended calculating a standardized effect size measures (e.g., Cohens' \(d\) Cohen, 1988) and the corresponding sampling variance. For example, after generating data for the two groups the mean difference can be standardized using the pooled standard deviation and applying the appropriate correction (Hedges, 1981; e.g., Hedges, 1989).

\color{black}

@@

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{set.seed}\NormalTok{(seed)}
\FunctionTok{sim\_study}\NormalTok{(}\AttributeTok{es =} \FloatTok{0.3}\NormalTok{, }\AttributeTok{nc =} \DecValTok{10000}\NormalTok{, }\AttributeTok{nt =} \DecValTok{10000}\NormalTok{, }\AttributeTok{aggregate =} \ConstantTok{FALSE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}
